Comparing vector-based and ACT-R memory models using large-scale datasets:\protect\\User-customized hashtag and tag prediction on Twitter and StackOverflow

### 2014-09-24 writeups/dissertation/dissertation.tex
foo

To better illustrate the dynamics of each user's tagging behavior,
profile plots for a single user for both StackOverflow and Twitter are included in Figure \ref{figPriorHTByTime}.

@@

The specific value of the hashtag used for both plots is a nominal variable and not important for this analysis.
What is interesting is the large tag space for these users.
There are bands of commonly-used tags for each user, but there are also many instances where new tags are used, and a large percentage of tags used are not in the most common bands.
For example, the StackOverflow user in Figure \ref{figPriorHTByTime} has a pronounced band where the same hashtag is chosen repeatedly over time. 
However there are many instances of other tags being used in parallel to the tag that is producing the band (that is repeatedly being used). 
The Twitter user has a few clusters where the same hashtag is chosen repeatedly but only for a brief period of time.
As time progresses, a band of choosing another hashtag is observed, and then the user switches to a new hashtag after that.
More recent posts for the Twitter user are more varied, and there is no longer a single hashtag that the user chooses repeatedly.
This helps illustrate how challenging modeling this task can be.
The model's goal is to predict at each instance what tag will be chosen, but to do so the model must predict the correct few tags from a much larger set of tags whose frequency of occurrence is changing over time.

@@

To identify the best decay rate value for the dataset slices,
each user's best decay rate value was taken and the results are plotted in histogram form in Figures \ref{figPriorSOQHistDsOL} and \ref{figPriorTwitterHistDsOL} respectively.

@@

The best-fit values are close to the \num{0.5} default value set in ACT-R when the optimized learning form is used.
Comparing across optimized learning and the standard form, the best-fit values for the standard form are slightly higher for both the StackOverflow and Twitter dataset slices.
Comparing best-fit values will be more thoroughly tested when looking at the values across all dataset slices.

@@

Aggregate best-fit decay rate values for each model are included in Figure \ref{figPriorDecay}.

@@

Decay rate values that produced the most accurate model performance are lower for the optimized learning form (\num{0.43}) of $B_{i}$ compared to the standard form (\num{0.80}).
The optimal values for optimized learning are near \num{0.5}, which lines up with default value used in ACT-R.
So the decay rate values for the standard form are slightly higher than the default values used in ACT-R, but the values found when optimized learning is used line up nicely with the ACT-R defaults.

@@

The impact of adding the offset to each of the models for StackOverflow and Twitter is shown in Figures \ref{figContextOffsetSO} and \ref{figContextOffsetT}.

\begin{figure}[!htbp]
  \scalebox{.6}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-OffsetSO-crop.pdf}}}
  \caption{
    Impact of adding offset component for StackOverflow.
    Error bars represent the 95\% bootstrapped confidence interval for the mean model accuracy across all runs in dataset.
    The random permutation models are represented as ``RP'' and Bayesian models as ``Bayes.''
}
  \label{figContextOffsetSO}
\end{figure}

\begin{figure}[!htbp]
  \scalebox{.6}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-OffsetT-crop.pdf}}}
  \caption{
    Impact of adding offset component for Twitter.
    All error bars are the 95\% bootstrapped confidence interval for the mean.
  }
  \label{figContextOffsetT}
\end{figure}

In both plots, if the model name includes ``full'' then all model components are included.
When they are enumerated, then at least one of the components is left out.
The random permutation and Bayesian models are plotted alongside each other so that comparisons across the two model types can be done easily.
So, for example, the ``standard prior model relaxed across posts without offset'' means that only the prior model component was used and the offset for that component was not subtracted off when determining activation.
As another example, the ``Bayes only orderless context'' means that only the context component of the Bayesian model was used (along with the offset), and that no order information was used for that contextual component term.

@@

@@

Model accuracy results for components of both the random permutation and Bayesian models grouped by Twitter popular-users subset are included in Figure \ref{figContextSubsets}.

@@

Although there is a main effect of subset, there are minimal to no practically significant interactions.
Relative model performance remains consistent across the four subsets.

@@

Since the interactions across popular-users datasets are minimal, model behavior for the popular-users subset for Twitter can be interpreted as behavior that will persist across different time periods,
if the same method of choosing popular hashtags is conducted at a later date.
It may also be the case that that model behavior does not depend on the specific method of choosing the popular hashtags, but since only a single method was used for this analysis,
the results from this study cannot by themselves be used to answer this question.

@@

Plots for the three methods of handling stop words for the random permutation model on the StackOverflow dataset are included in Figure \ref{figContextStopRPSO}.

@@

The results show that model accuracy improves if stop words are handled in any of the three ways (predefined stoplist filtering, entropy weighting, and frequency filtering).
Using a predefined stoplist makes the smallest improvement (\num{.45} to \num{.49}), while the entropy (\num{.60}) and frequency (\num{.63}) techniques improve performance the most,
and frequency shows a slight edge over entropy.
However, using the frequency-weighting technique produces more noise in accuracy measurements compared to the other techniques, at least when tested on words in the title of StackOverflow posts.

The three stop word techniques were then tested for the random permutation model on the Twitter dataset, and the results are included in Figure \ref{figContextStopRPT}.

@@

For the random permutation model for Twitter, the entropy weighting technique is the best method for handling stop words (\num{.25} to \num{.26}).
Removing stop-words based on a predetermined list actually decreases accuracy (\num{.25} to \num{.23}), and using data-driven frequency filtering has only a small to no effect (\num{.25} to \num{.25}).

@@

Similar results are found when the Bayesian model is tested on the Twitter dataset.

@@

Figure \ref{figContextStopBT} shows that entropy weighting is again the best performer (\num{.29} to \num{.32}), and frequency filtering decreases performance relative to no filtering (\num{.29} to \num{.28}).

@@

Model accuracy as a function of co-occurrence matrix size for StackOverflow is included in Figure \ref{figContextDocumentSizeSO}.

@@

Accuracy improves for both the Bayesian and random permutation models as the size of the documents used to generate the co-occurrence matrix increases.
However, model accuracy using the entropy weighting technique plateaus and slightly decreases for the largest document size for the random permutation model, while the frequency technique only begins to plateau. 
The Bayesian model does not plateau and accuracy continues to improve as document size increases.

@@

The overall trend of increasing accuracy with document size is present for Twitter as well.

@@

Both entropy weighting and frequency filtering in Figure \ref{figContextDocumentSizeT} plateau for the random permutation model with increasing document size.
Once again, the Bayesian model continues to improve with document size.

@@

Results for using different levels of compression for the random permutation model for the StackOverflow dataset are included in Figure \ref{figContextDimSO}.

@@

In the figure, the model name that does not have the number of rows in the matrix specified is the \num{2048}-row matrix.
This was done so that names in all of the plots would remain consistent across all plots, and the \num{2048}-row matrix is the size used for the random permutation model for all other results.
The plot is also broken out for different size datasets used to build the co-occurrence matrix.

@@

Model accuracy results after using the log-odds transformation are included in Figures \ref{figContextLogoddsSO} and \ref{figContextLogoddsT}.

@@

Using the log-odds transformation technique has a small effect on model accuracy for StackOverflow (\num{.45} to \num{.47}) and minimal to no effect on Twitter (\num{.25} to \num{.25}).
When the technique is used and accuracy is examined for a model with only a single predictor, the transformation does not change accuracy.
When predictors are combined and the full model is examined, the transformation is in the direction of slightly increasing accuracy, at least for the StackOverflow dataset.
However, any improvement when using the transformation is minimal.

@@

Model accuracy results after incorporating word order into the random permutation model are included in Figure \ref{figContextOrder}.

@@

All bar plots that include an order component where ``window'' or ``direction'' is not included in the title use the standard ordering method (no direction or windowing, all position information is included) 
This is the default ordering method used, so the names for these bar plots were kept consistent to the names used in other figures so that comparisons could be made across figures if necessary
(similar to what was done for the default row size for the random permutation model).
Base models that do not use entropy weighting were also included so that effect sizes could be compared between entropy weighting and adding word order information.

@@

Coefficients for each model component of the full models are included in Figures \ref{figContextCoeffSO} and \ref{figContextCoeffT}
Error bars represent the standard deviation of coefficient values for each component.

@@

The standard deviation is used instead of the confidence interval of the mean so that the plots show the overall stability of each coefficient across the different runs.
The range of the distribution of coefficient values within a component is much more narrow than the range of the mean coefficient values between the components of each model.
This shows that the model is not being overfit, as the weights for each component are relatively stable across the different runs.

@@

The results are included in Figures \ref{figContextPriorneededSO} and \ref{figContextPriorneededT}.

@@

For StackOverflow, accuracy remains the same when the prior component is removed for both the Bayesian (\num{.63} to \num{.63}) and random permutation (\num{.63} to \num{.63}) models.
Accuracy does dramatically decrease for Twitter when the prior is removed, however (\num{.32} to \num{.14} for Bayesian, \num{.25} to \num{.12} for random permutation).

@@

The results comparing the full model to a context-only only model for the StackOverflow popular-users dataset are included in Figure \ref{figPUAddPrior}.

@@

In this case, past tag history does contain unique variance for accurately predicting tag use.
Accuracy improves for both the Bayesian (\num{.51} to \num{.63}) and random permutation (\num{.47} to \num{.58}) models when the prior model component is combined with context compared to context alone.

@@

Model accuracy when using a user-customized co-occurrence matrix for StackOverflow and Twitter is shown in Figures \ref{figPUCustomSO} and \ref{figPUCustomT}.

@@

The results compare accuracy for models that use a global co-occurrence and user-specific co-occurrence matrix.
For StackOverflow, model accuracy for the user-specific co-occurrence matrix (labeled as ``user sji'') can be compared against the default global co-occurrence matrix.
Accuracy decreases dramatically when a user-specific co-occurrence matrix is used for both the Bayesian (\num{.64} to \num{.35}) and random permutation (\num{.58} to \num{.36}) models.

>>>>

To better illustrate the dynamics of each user's tagging behavior,
profile plots for a single user for both StackOverflow and Twitter are included in Figure \ref{figPriorHTByTime}.
The specific value of the hashtag used for both plots is a nominal variable and not important for this analysis.
What is interesting is the large tag space for these users.
There are bands of commonly-used tags for each user, but there are also many instances where new tags are used, and a large percentage of tags used are not in the most common bands.
For example, the StackOverflow user in Figure \ref{figPriorHTByTime} has a pronounced band where the same hashtag is chosen repeatedly over time. 
However there are many instances of other tags being used in parallel to the tag that is producing the band (that is repeatedly being used). 
The Twitter user has a few clusters where the same hashtag is chosen repeatedly but only for a brief period of time.
As time progresses, a band of choosing another hashtag is observed, and then the user switches to a new hashtag after that.
More recent posts for the Twitter user are more varied, and there is no longer a single hashtag that the user chooses repeatedly.
This helps illustrate how challenging modeling this task can be.
The model's goal is to predict at each instance what tag will be chosen, but to do so the model must predict the correct few tags from a much larger set of tags whose frequency of occurrence is changing over time.

@@

@@

To identify the best decay rate value for the dataset slices,
each user's best decay rate value was taken and the results are plotted in histogram form in Figures \ref{figPriorSOQHistDsOL} and \ref{figPriorTwitterHistDsOL} respectively.
The best-fit values are close to the \num{0.5} default value set in ACT-R when the optimized learning form is used.
Comparing across optimized learning and the standard form, the best-fit values for the standard form are slightly higher for both the StackOverflow and Twitter dataset slices.
Comparing best-fit values will be more thoroughly tested when looking at the values across all dataset slices.

@@

@@

Aggregate best-fit decay rate values for each model are included in Figure \ref{figPriorDecay}.
Decay rate values that produced the most accurate model performance are lower for the optimized learning form (\num{0.43}) of $B_{i}$ compared to the standard form (\num{0.80}).
The optimal values for optimized learning are near \num{0.5}, which lines up with default value used in ACT-R.
So the decay rate values for the standard form are slightly higher than the default values used in ACT-R, but the values found when optimized learning is used line up nicely with the ACT-R defaults.

@@

@@

The impact of adding the offset to each of the models for StackOverflow and Twitter is shown in Figures \ref{figContextOffsetSO} and \ref{figContextOffsetT}.
In both plots, if the model name includes ``full'' then all model components are included.
When they are enumerated, then at least one of the components is left out.
The random permutation and Bayesian models are plotted alongside each other so that comparisons across the two model types can be done easily.
So, for example, the ``standard prior model relaxed across posts without offset'' means that only the prior model component was used and the offset for that component was not subtracted off when determining activation.
As another example, the ``Bayes only orderless context'' means that only the context component of the Bayesian model was used (along with the offset), and that no order information was used for that contextual component term.

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-OffsetSO-crop.pdf}}}
  \caption{
    Impact of adding offset component for StackOverflow.
    Error bars represent the 95\% bootstrapped confidence interval for the mean model accuracy across all runs in dataset.
    The random permutation models are represented as ``RP'' and Bayesian models as ``Bayes.''
}
  \label{figContextOffsetSO}
\end{figure}

\begin{figure}[!htbp]
  \scalebox{.8}{\resizebox{\linewidth}{!}{\includegraphics{compareMeanDV-acc-OffsetT-crop.pdf}}}
  \caption{
    Impact of adding offset component for Twitter.
    All error bars are the 95\% bootstrapped confidence interval for the mean.
  }
  \label{figContextOffsetT}
\end{figure}

@@

In order to ensure that model results are not dependent on the specific Twitter subset collected,
it is important that the overall accuracy differences between the different models do not drastically shift from subset to subset.
That is, it is important that the effect size of model is much greater than the effect size of the interaction between Twitter subset and model.
A small interaction relative to the main effect means that the rank ordering of the different models remains consistent across the different Twitter subsets.
Further, a small effect size of subset relative to the effect of model means that the exact accuracy for each model remains consistent across the subsets.
To measure the effect sizes of subset, model type, and the interaction, a two-way between-subjects ANOVA was performed.

@@

Model accuracy results for components of both the random permutation and Bayesian models grouped by Twitter popular-users subset are included in Figure \ref{figContextSubsets}.
Accuracy is affected significantly by subset ($F(3, 340) = 764, p < .01$)), model ($F(4, 340) = 3617, p < .01$), and the interaction between model and subset ($F(12, 340) = 142, p < .01$).
However, the effect size of model type on accuracy is \num{8.6} times larger than the interaction term ($\eta ^ {2} = 0.770$ versus $\eta ^ {2} = 0.090$).
This shows that the performance of a model relative to the other models remains consistent across the four subsets.
Further, the effect size of model type is \num{6.3} times larger than the effect of Twitter subset ($\eta ^ {2} = 0.770$ versus $\eta ^ {2} = 0.122$).
This shows that not only is relative model performance maintained across Twitter subsets, but the exact value of model accuracy for each model remains reasonably consistent across subsets.

@@

@@

Since the effect sizes of subset and the interaction between model type and subset are minimal relative to the effect size of model,
model behavior for the popular-users subset for Twitter can be interpreted as behavior that will persist across different time periods,
if the same method of choosing popular hashtags is conducted at a later date.
It may also be the case that that model behavior does not depend on the specific method of choosing the popular hashtags, but since only a single method was used for this analysis,
the results from this study cannot by themselves be used to answer this question.

@@

Plots for the three methods of handling stop words for the random permutation model on the StackOverflow dataset are included in Figure \ref{figContextStopRPSO}.
The results show that model accuracy improves if stop words are handled in any of the three ways (predefined stoplist filtering, entropy weighting, and frequency filtering).
Using a predefined stoplist makes the smallest improvement (\num{.45} to \num{.49}), while the entropy (\num{.60}) and frequency (\num{.63}) techniques improve performance the most,
and frequency shows a slight edge over entropy.
However, using the frequency-weighting technique produces more noise in accuracy measurements compared to the other techniques, at least when tested on words in the title of StackOverflow posts.

@@

The three stop word techniques were then tested for the random permutation model on the Twitter dataset, and the results are included in Figure \ref{figContextStopRPT}.
For the random permutation model for Twitter, the entropy weighting technique is the best method for handling stop words (\num{.25} to \num{.26}).
Removing stop-words based on a predetermined list actually decreases accuracy (\num{.25} to \num{.23}), and using data-driven frequency filtering has only a small to no effect (\num{.25} to \num{.25}).

@@

@@

Similar results are found when the Bayesian model is tested on the Twitter dataset.
Figure \ref{figContextStopBT} shows that entropy weighting is again the best performer (\num{.29} to \num{.32}), and frequency filtering decreases performance relative to no filtering (\num{.29} to \num{.28}).

@@

@@

Model accuracy as a function of co-occurrence matrix size for StackOverflow is included in Figure \ref{figContextDocumentSizeSO}.
Accuracy improves for both the Bayesian and random permutation models as the size of the documents used to generate the co-occurrence matrix increases.
However, model accuracy using the entropy weighting technique plateaus and slightly decreases for the largest document size for the random permutation model, while the frequency technique only begins to plateau. 
The Bayesian model does not plateau and accuracy continues to improve as document size increases.

@@

@@

The overall trend of increasing accuracy with document size is present for Twitter as well.
Both entropy weighting and frequency filtering in Figure \ref{figContextDocumentSizeT} plateau for the random permutation model with increasing document size.
Once again, the Bayesian model continues to improve with document size.

@@

@@

Results for using different levels of compression for the random permutation model for the StackOverflow dataset are included in Figure \ref{figContextDimSO}.
In the figure, the model name that does not have the number of rows in the matrix specified is the \num{2048}-row matrix.
This was done so that names in all of the plots would remain consistent across all plots, and the \num{2048}-row matrix is the size used for the random permutation model for all other results.
The plot is also broken out for different size datasets used to build the co-occurrence matrix.

@@

@@

Model accuracy results after using the log-odds transformation are included in Figures \ref{figContextLogoddsSO} and \ref{figContextLogoddsT}.
Using the log-odds transformation technique has a small effect on model accuracy for StackOverflow (\num{.45} to \num{.47}) and minimal to no effect on Twitter (\num{.25} to \num{.25}).
When the technique is used and accuracy is examined for a model with only a single predictor, the transformation does not change accuracy.
When predictors are combined and the full model is examined, the transformation is in the direction of slightly increasing accuracy, at least for the StackOverflow dataset.
However, any improvement when using the transformation is minimal.

@@

@@

Model accuracy results after incorporating word order into the random permutation model are included in Figure \ref{figContextOrder}.
All bar plots that include an order component where ``window'' or ``direction'' is not included in the title use the standard ordering method (no direction or windowing, all position information is included) 
This is the default ordering method used, so the names for these bar plots were kept consistent to the names used in other figures so that comparisons could be made across figures if necessary
(similar to what was done for the default row size for the random permutation model).
Base models that do not use entropy weighting were also included so that effect sizes could be compared between entropy weighting and adding word order information.

@@

@@

Coefficients for each model component of the full models are included in Figures \ref{figContextCoeffSO} and \ref{figContextCoeffT}
Error bars represent the standard deviation of coefficient values for each component.
The standard deviation is used instead of the confidence interval of the mean so that the plots show the overall stability of each coefficient across the different runs.
The range of the distribution of coefficient values within a component is much more narrow than the range of the mean coefficient values between the components of each model.
This shows that the model is not being overfit, as the weights for each component are relatively stable across the different runs.

@@

@@

The results are included in Figures \ref{figContextPriorneededSO} and \ref{figContextPriorneededT}.
For StackOverflow, accuracy remains the same when the prior component is removed for both the Bayesian (\num{.63} to \num{.63}) and random permutation (\num{.63} to \num{.63}) models.
Accuracy does dramatically decrease for Twitter when the prior is removed, however (\num{.32} to \num{.14} for Bayesian, \num{.25} to \num{.12} for random permutation).

@@

@@

The results comparing the full model to a context-only only model for the StackOverflow popular-users dataset are included in Figure \ref{figPUAddPrior}.
In this case, past tag history does contain unique variance for accurately predicting tag use.
Accuracy improves for both the Bayesian (\num{.51} to \num{.63}) and random permutation (\num{.47} to \num{.58}) models when the prior model component is combined with context compared to context alone.

@@

@@

Model accuracy when using a user-customized co-occurrence matrix for StackOverflow and Twitter is shown in Figures \ref{figPUCustomSO} and \ref{figPUCustomT}.
The results compare accuracy for models that use a global co-occurrence and user-specific co-occurrence matrix.
For StackOverflow, model accuracy for the user-specific co-occurrence matrix (labeled as ``user sji'') can be compared against the default global co-occurrence matrix.
Accuracy decreases dramatically when a user-specific co-occurrence matrix is used for both the Bayesian (\num{.64} to \num{.35}) and random permutation (\num{.58} to \num{.36}) models.

@@
